---
title: "AnimalTrackR Python Dependencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AnimalTrackR Python Dependencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(AnimalTrackR)
```

AnimalTrackR has several python dependencies as YOLO is a python-based object detection framework. The vast majority of users will not need to interact directly with these dependencies. If you have any doubt, you go to the function documentation for `create_TrackR_env()`. This will install dependencies and handle environment creation correctly for almost all users.

If you have specific requirements based on your system, or other analysis pipelines, you can use this guide to create a custom environment that has the correct python version and packages to run AnimalTrackR. This guide assumes you have a working knowledge of Anaconda/Miniconda and managing python enviornments and packages.

## Environment

AnimalTrackR relies on a conda environment. It will not work with other types of python enviornment (e.g. virtual environments). Therefore, if you wish to create a custom environment you must first install conda. You can install the full release of [Anaconda](https://www.anaconda.com/docs/getting-started/anaconda/install), but we recommend the lightweight installation of [Miniconda](https://www.anaconda.com/docs/getting-started/miniconda/install). AnimalTrackR uses Miniconda by default.

## Python version

AnimalTrackR has been tested using python 3.11. Many of its dependencies will work with other python versions, but some issus have been reported with dependency conflicts in a number of commputer vision libraries, particularly pytorch, that may become problematic. Users can experiment with other python versions if they need to but we strongly reccomend staying with version 3.11.

## Python packages

AnimalTrackR has two core python package dependencies; ultalytics and opencv.

**ultralytics** is a python API for YOLO. More details are available [here](https://docs.ultralytics.com/)
**opencv** is a widely used computer vision library that is essential for basic operations performed by AnimaltrackR. See [here](https://pypi.org/project/opencv-python/) for details.

If you are using a CPU only installing the most recent version of these packages should not give you any issues.

## Python packages for GPU support

If you are using an NVIDIA gpu there are some additional dependencies you will need. Where package versions are provided in this list they are **essential** to avoid conflicts:

* pytorch==2.1
* torchvision
* pytorch-cuda (version must correspond to cuda version, which can be found by running `nvidia-smi` in a command prompt)
* numpy<2 (numpy version must be less than 2 to avoid conflicts)
